---
title: "The genius invention about aesthetics in ggplot2"
subtitle: "The common and uncommon usages of aes() and its implementation"
description: "A personal appreciation of aes() in ggplot2: from basic mappings to advanced delayed evaluation, 'tidy' programming patterns, and seamless tidyverse integration."
date: "2026-01-03"
categories: [R, ggplot2, visualization, tidyverse]
format:
    html:
        toc: true
        toc-float: true
        toc-depth: 3
        number-sections: true
        code-fold: false
        code-tools: false
        theme: default
        fig-width: 10
        fig-height: 6
        fig-cap-location: bottom
        code-annotations: hover
execute:
    echo: true
    warning: false
    message: false
filters:
    - social-share
    - collapse-output
share:
    permalink: "https://joshuamarie.com/posts/14-aes-ggplot2"
    description: "Aesthetics in ggplot2 – Joshua Marie"
    twitter: true
    facebook: true
    reddit: true
    stumble: true
    tumblr: true
    linkedin: true
    email: true
    mastodon: true
    bsky: true
    location: "before-body"
engine: knitr
---

# What do I love about _aes()_?

The `aes()` function is a **domain-specific language** (DSL) embedded inside R that feels almost magical thanks to lazy evaluation, quosures, and deep integration with the `{tidyverse}`.

What keeps pulling me back is how elegantly it balances simplicity for beginners with incredible power for advanced use — all without ever feeling clunky.

# Mapped vs Fixed Aesthetics

One of the first things to understand is the distinction between **mapped** and **fixed** aesthetics.

```{r}
box::use(
    ggplot2[ggplot, aes, geom_point]
)
```

::::: columns
::: {.column width="50%"}
## The colors are mapped through groups

```{r}
ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
    geom_point()
```
:::

::: {.column width="50%"}
## The color is fixed for all points

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) +
    geom_point(color = "blue")
```
:::
:::::

When you put an aesthetic inside `aes()`, it's **mapped** to a variable. Outside `aes()`, it's **fixed** to a constant value. This distinction is crucial but often trips up beginners.

# The _group_ Aesthetic

The `group` aesthetic controls how `{ggplot2}` groups data for statistical transformations and geoms like lines.

```{r}
box::use(
    ggplot2[ggplot, aes, geom_line]
)
```

::::: columns
::: {.column width="50%"}
## Without group, one line connects all points

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) +
    geom_line()
```
:::

::: {.column width="50%"}
## With group, separate lines for each cylinder count

```{r}
ggplot(mtcars, aes(x = mpg, y = hp, group = cyl)) +
    geom_line()
```
:::
:::::

Often, `group` is set automatically when you use `color`, `linetype`, or `fill`, but sometimes you need explicit control.

# Stage-Based Transformations: _after_stat()_ and _after_scale()_

I already made a blog talking about delaying `aes()` arguments, you can check it [here](https://joshuamarie.com/posts/09-ggplot2/#the-art-of-delay). These functions let you compute aesthetics based on statistical transformations or after scale transformations.

```{r}
box::use(
    ggplot2[
        ggplot, aes, geom_histogram, geom_point, 
        after_stat, after_scale, alpha
    ]
)
```

1.  Color bars based on computed count from `stat_bin`

    ```{r}
    ggplot(mtcars, aes(x = mpg)) +
        geom_histogram(aes(fill = after_stat(count)), bins = 10)
    ```

2.  Use `after_scale()` to modify colors after scale mapping

    ```{r}
    ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
        geom_point(aes(color = after_scale(alpha(color, 0.5))), size = 4)
    ```

`after_stat()` accesses variables computed by statistical transformations (like `count`, `density`, etc.). `after_scale()` works with the final scaled aesthetic values, allowing post-processing of colors, sizes, and more.

# Tidy Evaluation and Programming with _aes()_

When you're writing functions that create `{ggplot2}` plots, you need to programmatically pass variable names. This is where tidy evaluation with `{{ }}` (curly-curly) and `!!` (bang-bang) comes in.

1.  The double curly brackets `{{ }}` is more recommended

    ```{r}
    plot_scatter = function(data, x_var, y_var, color_var) {
        box::use(
            ggplot2[
                ggplot, aes, geom_point, theme_minimal
            ]
        )
        
        ggplot(data, aes(x = {{ x_var }}, y = {{ y_var }}, color = {{ color_var }})) +
            geom_point() +
            theme_minimal()
    }

    plot_scatter(mtcars, mpg, hp, cyl)
    ```

2.  With `rlang::ensym()`, it accepts both string and a bare column

    ```{r}
    plot_scatter_string = function(data, x_var, y_var) {
        box::use(
            ggplot2[ggplot, aes, geom_point],
            rlang[ensym]
        )
        
        ggplot(data, aes(x = !!ensym(x_var), y = !!ensym(y_var))) +
            geom_point()
    }

    plot_scatter_string(mtcars, "mpg", "hp")
    # plot_scatter_string(mtcars, "mpg", hp)
    # plot_scatter_string(mtcars, mpg, "hp")
    # plot_scatter_string(mtcars, mpg, hp)
    ```

This works because `aes()` uses quosures (quoted expressions with their environment) internally, which are part of `{rlang}`'s non-standard evaluation framework.

# Integration with the Tidyverse {#dplyr-ggplot2-pipe}

The `aes()` function plays beautifully with `{tidyverse}` workflows because they share the same evaluation model.

```{r}
#| label: "dplyr-ggplot2-pipe"
box::use(
    dplyr[mutate, case_when], 
    ggplot2[ggplot, aes, geom_point, labs, theme_minimal]
)

mtcars |> 
    mutate(
        efficiency = mpg / hp,
        cyl_category = case_when(
            cyl == 4 ~ "Small",
            cyl == 6 ~ "Medium",
            cyl == 8 ~ "Large"
        )
    ) |> 
    ggplot() +
    aes(x = wt, y = efficiency, color = cyl_category) + 
    geom_point(size = 3) +
    labs(title = "Efficiency by Weight and Engine Size") + 
    theme_minimal()
```

You can pipe data through `{dplyr}` transformations and directly into `{ggplot2}`, and all the column names work seamlessly. This should be trivial to you since this is more taught by most `{ggplot2}` tutorials. 

# Other aesthetic 'weaponries' you want to use

Beyond the common patterns, `aes()` has several advanced capabilities that showcase its flexibility. 

Here, I can name few: 

## Inheriting and Overriding Aesthetics

You can set aesthetics at the plot level and override them in specific layers. This is useful for creating complex plots with different mappings per layer.

```{r}
box::use(
    ggplot2[
        ggplot, aes, geom_point, geom_smooth, labs, 
        geom_line, theme_minimal
    ]
)

ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
    geom_point(size = 3) +
    # Override color for this layer only
    geom_smooth(
        aes(color = NULL), 
        method = "lm", 
        color = "#061E29", 
        se = FALSE,
        linetype = "dashed"
    ) + 
    # Then the regression lines for each group
    geom_line(stat = "smooth", method = "lm", se = FALSE, alpha = 0.5) + 
    labs(title = "Individual trends vs overall trend") + 
    theme_minimal()
```

## Using _aes()_ with Statistical Transformations

Some geometries perform statistical transformations and create new variables. You can map aesthetics to these computed variables using `after_stat()` or by knowing the variable names.

```{r}
box::use(
    ggplot2[
        ggplot, aes, geom_histogram, geom_line, after_stat, 
        scale_fill_viridis_c, scale_color_viridis_c, 
        labs, theme_minimal, theme, margin, 
        element_text
    ]
)

ggplot(mtcars, aes(x = mpg)) +
    geom_histogram(
        aes(
            y = after_stat(density),  
            fill = after_stat(density > 0.05)  
        ),
        bins = 30,
        color = "white",
        linewidth = 0.2
    ) +
    geom_line(
        stat = "density", 
        mapping = aes(
            y = after_stat(density),  
            color = after_stat(density > 0.05)  
        ), 
        show.legend = FALSE
    ) + 
    scale_fill_viridis_c(option = "plasma") + 
    scale_color_viridis_c(option = "plasma") + 
    labs(
        title = "Histogram with after_stat()",
        subtitle = "Y-axis shows density, fill color shows count",
        x = "Miles per gallon",
        y = "Density",
        fill = "Count"
    ) +
    theme_minimal(base_size = 14) +
    theme(
        plot.title = element_text(face = "bold"),
        legend.position = "right",
        plot.margin = margin(15, 15, 15, 15)
    )
```

Here we're using `after_stat(scaled)` to access the scaled density values and conditionally coloring regions.

## Constant aesthetics inside _aes()_

In this [section](#dplyr-ggplot2-pipe), we piped down the result from the data wrangling with `{dplyr}` into `{ggplot2}` layers. How about the transformation happened within the layer itself? Reminder that `aes()` is not far from Lisp macros and handles homoiconicity, which means it will handle an expression

```{r}
box::use(
    ggplot2[ggplot, aes, geom_point, labs, scale_color_identity],
    dplyr[if_else]
)

ggplot(mtcars) + 
    aes(x = wt, y = mpg, color = I(if_else(mpg > 20, "darkgreen", "darkred"))) + 
    geom_point(size = 4) +
    scale_color_identity(
        guide = "legend", 
        labels = c("Low MPG", "High MPG"),
        breaks = c("darkred", "darkgreen")
    ) +
    labs(color = "Efficiency")
```

`I()` tells `{ggplot2}` to use the values as-is without scaling, while `scale_color_identity()` creates a legend.

## Dynamic Aesthetics

You can create aesthetics on-the-fly using R functions directly in `aes()`. Here, I _cut_ down `hp` into 3 groups, then use it as a basis for the shapes. 

```{r}
box::use(
    ggplot2[ggplot, aes, geom_point, labs, theme_minimal]
)

ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point(
        aes(
            size = cyl,
            shape = cut(
                hp, 
                breaks = c(0, 100, 200, 400), 
                labels = c("Low", "Medium", "High")
            )
        ), 
        alpha = 0.7
    ) +
    labs(size = "Horsepower", shape = "Cylinders") +
    theme_minimal()
```

# My argument regarding ggplot2 implementation

I've argued with Python users who say "there's `{plotnine}` in Python and you have all the features in `{ggplot2}` and you can't look back." — it's not even close. I maintain that it's just a pale imitation: you can't really fully replicate this package. It'll take a lot to FULLY replicate it, and it's TOO complicated for Python to replicate.

The thing is, R in general is functional; the arguments are *lazily evaluated* by default and similar to Lisp macros, where the AST can be transformed/modified during compile-time—you can pretty much manipulate and inspect expressions before they're evaluated, giving you metaprogramming capabilities that are first-class citizens in R.

## Comparison

Here's a side by side comparison:

::: panel-tabset
### In R with ggplot2

This package uses first-class metaprogramming — you can use clean, bare variable names. You can also override and transform inline.

```{r}
box::use(
    ggplot2[ggplot, aes, geom_point, geom_smooth]
)

ggplot(mtcars, aes(x = mpg, y = hp * 2, color = factor(cyl))) +
    geom_point() +
    geom_smooth(aes(y = hp * 2), method = "lm")
```

### In Python with plotnine

Python in general lacks this environment as in R, we refer it as ["computing on the language"](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Computing-on-the-language-1), which dubbed later by Hadley Wickham as "non-standard evaluation", one of the techniques used in [R metaprogramming](https://adv-r.hadley.nz/metaprogramming.html).

```{python}
#| include: false
#| echo: false
#| warning: false
#| message: false
mtcars = r.mtcars
```

Supplying arguments within `aes()` from `{plotnine}` package works with strings. Yes, it works, but notice the quotes.

```{python}
from plotnine import ggplot, aes, geom_point, geom_smooth

(ggplot(mtcars, aes(x = 'mpg', y = 'hp * 2', color = 'factor(cyl)')) +
    geom_point() +
    geom_smooth(aes(y = 'hp * 2'), method = 'lm')) 
```
:::

## Why stringly typed code is fundamentally problematic

Let me clarify: The Python version suffers from being [stringly typed](http://wiki.c2.com/?StringlyTyped). The fact that `{plotnine}` forces to approximate things in `{ggplot2}` R using strings where structured data types should be used. This creates concrete technical problems, so hear me out:

1.  Loss of type safety and compile-time guarantees. All of the arguments supplied into `aes()` are valid strings to Python, but only one works

    ``` python
    aes(x = 'mpg')      # 1 
    aes(x = 'mpgg')     # 2 
    aes(x = 'mpg + ')   # 3 
    aes(x = 42)         # 4 
    ```

    1.  This syntax is correct
    2.  You made a typo — no error until runtime
    3.  Syntax error in the mini-language - no error until runtime
    4.  Wrong type, maybe? Depends on plotnine's implementation

    In R, `aes(x = mpg)` is a proper expression that participates in R's type system and evaluation model.

2.  You're implementing a parser for a mini-language

    Plotnine must parse strings like `'np.log(mpg) if mpg > 0 else 0'` at runtime. This means:

    -   Writing a lexer/parser for Python expressions
    -   Handling operator precedence, function calls, conditionals
    -   Maintaining this parser as Python evolves
    -   Debugging when the parser behaves differently than Python itself

    *Note that supplying an argument with some kind of transformation, e.g. `aes(y = 'np.log(mpg)')` still works.*

    In R, it handles (lazy) evaluation:

    ``` r
    aes(x = ifelse(mpg > 0, log(mpg), 0)) 
    ```

    You can take this in different level with `stage()`, `after_stat()`, and `after_scale()`.

<!-- # Why This Implementation Matters -->

<!-- This design gives `{ggplot2}` its declarative, composable nature. You're building a specification of what you want, and the system figures out the rendering. Python's eager evaluation makes this pattern fundamentally harder to implement cleanly—you'd need to build an entire mini-language or use metaclasses and descriptors to approximate this behavior. -->

<!-- That's why `{plotnine}` always feels like it's playing catch-up: it's trying to replicate behavior that's deeply tied to R's evaluation model and metaprogramming capabilities. -->

# Resources

-  [ggplot2 Official Documentation](https://ggplot2.tidyverse.org/)
-  [ggplot2 Aesthetic specification](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html)
